---
title: "Bellabeat - Case Study 2"
author: "Muhammad Farjad Akmal"
date: "December 19, 2021"
output: 
  html_document:
    toc: True
    number_sections: True
    theme: "journal"
---
## Credits
A very special thanks to Julen Aranguren, a Kaggler, whose work was adopted in the creation of this notebook. Kindly check his profile and his notebook at: https://www.kaggle.com/julenaranguren/bellabeat-case-study/notebook

## Business Task
Analysing device fitness data to unlock new growth opportunities for the company. 

## Objective
To analyse smart device usage data in order to gain insight into how consumers use non-Bellabeat smart devices.

## Loading Libraries - required for the data visualization, data wrangling and data formmatting.
```{r}
library(tidyverse)
library(reshape2)
library(scales)
```
## Phase 2 : Data Preparation 


### Loading CSV files
```{r}
daily_Activity <- read.csv("../Fitabase_Data/dailyActivity_merged.csv")
sleep_day <- read.csv("../Fitabase_Data/sleepDay_merged.csv")
dailyCalories <- read.csv("../Fitabase_Data/dailyCalories_merged.csv")
dailyIntensity <- read.csv("../Fitabase_Data/dailyIntensities_merged.csv")
dailySteps <- read.csv("../Fitabase_Data/dailySteps_merged.csv")
weightLogInfo <- read_csv("../Fitabase_Data/weightLogInfo_merged.csv")
```

### Exploring a few key tables

Take a look at the daily_activity data.
```{r}
head(daily_Activity)
```
Take a look at the sleep_Day data.
```{r}
head(sleep_day)
```

Take a look at the dailyCalories data.
```{r}
head(dailyCalories)
```

Take a look at the dailyIntensity data.
```{r}
head(dailyIntensity)
```

Take a look at the dailySteps data.
```{r}
head(dailySteps)
```

Take a look at the weightLogInfo data.
```{r}
head(weightLogInfo)
```


## Data Merging
Merge all the daily tables data and the sleep data. Join with ID and date columns.

```{r}
merge_daily_1 <- merge(dailyCalories, daily_Activity, by = c("Id", "Calories"))
merge_daily_2 <- merge(dailyIntensity, dailyIntensity, by = c("Id", "ActivityDay", "SedentaryMinutes", "LightlyActiveMinutes", "FairlyActiveMinutes", "VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance","ModeratelyActiveDistance","VeryActiveDistance"))
                       
merge_daily <- merge(merge_daily_1, merge_daily_2, by = c("Id", "ActivityDay", "SedentaryMinutes", "LightlyActiveMinutes", "FairlyActiveMinutes", "VeryActiveMinutes", "SedentaryActiveDistance", "LightActiveDistance","ModeratelyActiveDistance","VeryActiveDistance")) %>% 
  select (-ActivityDay) %>% 
  rename(Date = ActivityDate)

daily_data <- merge(merge_daily, sleep_day, by = "Id", all = TRUE) %>% 
  drop_na() %>% 
  select(-SleepDay, -TrackerDistance)

options(repr.plot.width=15, repr.plot.height=8)
```

The first few rows of daily_data.
```{r}
head(daily_data)
```

Identify all the columns in the daily_data.
```{r}
colnames(daily_data)
```

Summary statistics of daily_data 
```{r}
summary(daily_data)
```

## Phase 3 & 4: Process and Analyse 
### Data cleaning and preparation for analysis
```{r}
data_by_usertype <- daily_data %>% 
  summarise(
    user_type = factor(case_when(
      SedentaryMinutes > mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Sedentary", 
      SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes > mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Lightly Active",
      SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes > mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Fairly Active",
      SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes > mean(VeryActiveMinutes) ~ "Very Active",
), levels = c("Sedentary", "Lightly Active", "Fairly Active", "Very Active")), Calories,  .group = Id) %>% 
  drop_na()
```


### Data Visualization and Analysis
#### Calories Burned w.r.t., User Type  
```{r}
data_by_usertype %>% 
  group_by(user_type) %>% 
  summarise(total = n()) %>% 
  mutate(totals = sum(total)) %>% 
  group_by(user_type) %>% 
  summarise(total_percent = total / totals) %>% 
  ggplot(aes(user_type, y = total_percent, fill = user_type)) + 
  geom_col()+
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "none") + 
  labs(title = "User type distribution", x =NULL) +
  theme(legend.position = "none", text = element_text(size = 15), plot.title = element_text(hjust = 0.5))

ggplot(data_by_usertype, aes(user_type, Calories, fill = user_type)) + 
  geom_boxplot() + 
  theme(legend.position = "none") +
  labs(title = "Calories burned by User type", x = NULL) + 
  theme(legend.position = "none", text = element_text(size = 15), plot.title = element_text(hjust = 0.5))
  
```


### Check how is TotalDistance/TotalSteps is related to Calories Burned
```{r}
daily_data %>%
summarise(
distance = factor(case_when(
    TotalDistance < 4.5 ~ "< 4.5 mi",
    TotalDistance >= 4.5 & TotalDistance <= 7 ~ "4.5 > & < 7 mi",
    TotalDistance > 7 ~ "> 7 mi",
),levels = c("> 7 mi","4.5 > & < 7 mi","< 4.5 mi")),
steps = factor(case_when(
    TotalSteps < 6000 ~ "< 6k steps",
    TotalSteps >= 6000 & TotalSteps <= 10000 ~ "6k > & < 10k Steps",
    TotalSteps > 10000 ~ "> 10k Steps",
),levels = c("> 10k Steps", "6k > & < 10k Steps", "< 6k steps")),
Calories) %>%
ggplot(aes(steps,Calories,fill=steps)) +
    geom_boxplot() +
    facet_wrap(~distance)+
    labs(title="Calories burned by Steps and Distance",x=NULL) +
    theme(legend.position = "none", axis.text.x = element_text(size = 10, angle = 45, hjust =1), plot.title = element_text(hjust = 0.5))
    
```

### Sleeping patterns and quality of sleep
```{r}
sleepType_by_userType <- daily_data %>%
group_by(Id) %>%
summarise(
user_type = factor(case_when(
    SedentaryMinutes > mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Sedentary",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes > mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Lightly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes > mean(FairlyActiveMinutes) & VeryActiveMinutes < mean(VeryActiveMinutes) ~ "Fairly Active",
    SedentaryMinutes < mean(SedentaryMinutes) & LightlyActiveMinutes < mean(LightlyActiveMinutes) & FairlyActiveMinutes < mean(FairlyActiveMinutes) & VeryActiveMinutes > mean(VeryActiveMinutes) ~ "Very Active",
),levels=c("Sedentary", "Lightly Active", "Fairly Active", "Very Active")),
sleep_type = factor(case_when(
    mean(TotalMinutesAsleep) < 360 ~ "Bad Sleep",
    mean(TotalMinutesAsleep) > 360 & mean(TotalMinutesAsleep) <= 480 ~ "Normal Sleep",
    mean(TotalMinutesAsleep) > 480 ~ "Over Sleep",
),levels=c("Bad Sleep", "Normal Sleep", "Over Sleep")), total_sleep = sum(TotalMinutesAsleep) ,.groups="drop"
) %>%
drop_na() %>%
group_by(user_type) %>%
summarise(bad_sleepers = sum(sleep_type == "Bad Sleep"), normal_sleepers = sum(sleep_type == "Normal Sleep"),over_sleepers = sum(sleep_type == "Over Sleep"),total=n(),.groups="drop") %>%
group_by(user_type) %>%
summarise(
    bad_sleepers = bad_sleepers / total, 
    normal_sleepers = normal_sleepers / total, 
    over_sleepers = over_sleepers / total,
    .groups="drop"
)
```

### Plotting the above findings
```{r}
sleepType_by_userType_melted<- melt(sleepType_by_userType, id.vars = "user_type")

ggplot(sleepType_by_userType_melted, aes(user_type, value, fill = variable)) +
geom_bar(position = "dodge", stat = "identity") +
scale_y_continuous(labels = scales::percent) +
labs(x=NULL, fill="Sleep type") + 
theme(legend.position="bottom",text = element_text(size = 15),plot.title = element_text(hjust = 0.5))
```

